Moral Foundations of U.S. Political News Organizations

### 2024-11-30 manuscript_version/revision/manuscript_metapsych_revision.Rmd
finalizing the revision letter and manuscript

abstract: |
  Partisan differences and diviseness have become an increasing hot topic in psychological research. Many theories have been proposed to explain these differences and divisions including Moral Foundations Theory. The current research seeks to use a linguistic measure of Moral Foundations, the Moral Foundations Dictionary (MFD), to test the theory in terms of predicted partisan differences. In Study 1, we explore the MFD-Moral Foundations Questionnaire's measurement overlap using a multi-trait multi-method model. These latent trait models were used to inform possible improvement of the MFD. Next, we examined two replication extensions of previous work exploring the differences in political articles on moral differences Through web scraping, we extracted articles from popular partisan news sources' websites, calculated MFD word frequencies, and identified words' respective valences. This process attempts to uncover news outlets' positive or negative endorsements of certain moral dimensions concomitant with a particular ideology. In Study 2, we gathered political articles from four sources. We were unable to reveal significant differences in moral endorsements, but we solidified the method to be employed in further research. In Study 3, we expanded their number of sources to 10 and analyzed articles that pertain to two specific topics: the 2018 confirmation hearings of U.S. Supreme Court Justice Brett Kavanaugh and the partial U.S. Government Shutdown of 2018-2019. Once again, no significant differences in moral endorsements were found. Together with past work, the results shed doubt on the validity of the MFD as a reliable measurement tool.

@@

The field of moral psychology has a long history with modern moral
psychology beginning in the late 1960s with Lawrence Kohlberg's theory
of moral development [@Kohlberg1977]. Since then, Kohlberg's theory has
been highly criticized with numerous theories and hypotheses proposed to
replace it. One of these theories is Moral Foundations Theory (MFT)
proposed by Haidt and colleagues [@haidt2004; @Haidt2007]. While the
theory itself has been widely criticized and debated, the goal of the
current work is to tackle a related measurement issue, the Moral
Foundations Dictionary [MFD, @Graham2009]. @Graham2009 developed the
MFD - a lexicon-based linguistic measure of MFT - arguing that moral
concerns and the propositions of MFT could be measured and tested with
the words people use. Here we first review evidence and criticisms
against this assumption. We then provide our own test of the MFD using
partisan news articles.

@@

Haidt and Joseph [-\@haidt2004] and Haidt and Graham [-@Haidt2007]
formulated MFT as a method by which to capture the entirety of humans'
moral domain. They argued that older theories of moral psychology were
focused primarily on issues of justice, fairness, and caring, which are
individually focused foundations of morality that align with the beliefs
of political liberals. In other words, moral psychology ignored the
valid moral foundations of conservatives. MFT holds that people's moral
domain can be mapped by quantifying their endorsement of five moral
foundations: *harm/care*, *fairness/reciprocity*, *ingroup/loyalty*,
*authority/respect*, and *purity/sanctity* [@Haidt2007]. Thus, they
proposed that the addition of these two additional domains could
adequately capture the broader spectrum of morality.

Measurement of MFT has traditionally focused on two options: 1) the
Moral Foundations Questionnaire (MFQ) - a typical Likert-type survey
instrument that measures moral relevance of each foundation to an
individual and the moral judgment of various scenarios depicting each
foundation, and 2) the Moral Foundations Dictionary (MFD) which was
designed to capture language's role in moral reasoning and justification
as used in speech and text [@Graham2009; @graham2011]. The MFQ is a
30-item scale with 15 items on the moral relevance and judgment
scenarios. The 15 moral relevance items are equally divided among the
five foundations and examine how relevant a condition is to making a
moral judgment on a scale of 1 (*not at all relevant*) to 6 (*extremely
relevant*). These relevance items include examples such as: "Whether or
not someone used violence (*harm*)," "Whether or not someone was denied
his or her rights (*fairness*)," "Whether or not someone showed a lack
of loyalty (*ingroup*)," "Whether or not an action caused chaos or
disorder (*authority*)," and "Whether or not someone did something
disgusting (*purity*)." The moral judgments items are also equally
divided between the foundations and ask on a six-point scale how much
one agrees with each of the statements. These judgment items include:
"One of the worst things a person can do is hurt a defenseless animal
(*harm*)," "Justice is the most important requirement of a society
(*fairness*)," "I am proud of my country's history (*ingroup*)," "Men
and women each have different roles to play in society (*authority*),"
and "Chastity is an important and valuable virtue (*purity*)."

The internal consistency of the MFQ from @Graham2011 was $\alpha$ = .73
averaged across subscales with a range of $\alpha$ = .65-.84. Across six
studies, the MFQ was found to have an average Cronbach's alpha of .63
for *harm*, .64 for *fairness*, .56 for *ingroup*, .59 for *authority*,
and .71 for *purity* [@Federico2013; @Graham2009; @Graham2012;
@Weber2013]. Test-retest reliability was *r* = .68-.82 using a sample of
123 college students. Confirmatory factor analysis supported a
well-fitted five-factor model (*harm/care*, *fairness/reciprocity*,
*ingroup/loyalty*, *authority/respect*, and *purity/sanctity*) over two,
individual (*harm* and *fairness*) versus group (*ingroup*, *authority*,
and *purity*) foundations, or three, autonomy (*harm*, *fairness*),
community (*ingroup*, *authority*), and divinity (*purity*) ethics,
foundations factor model. The five-factor structure also fit for
non-Western samples, thus, providing evidence of the MFQ
generalizability. Convergent validity was supported with correlations on
other measures of morality [@Graham2011].

The MFD is composed of 259 words, with around 50 words assigned to each
of the five foundations. @Graham2009 created a preliminary list of words
that they believed would be associated with the five foundations. Words
such as *war* and *peace* should indicate a greater concern with *harm*
foundation whereas words such as *homeland* and *terrorism* should
indicate a greater concern with the *ingroup* foundation. The other
foundation dictionaries include *equal* and *justice* (*fairness*),
*honor* and *protest* (*authority*), and *holy* and *sin* (*purity*). To
validate the word sets, @Graham2009 examined the frequency of MFD words
in liberal and conservative sermons using the Linguistic Inquiry and
Word Count [LIWC; @Pennebaker2007] computer program. They found liberal
ministers used *harm*, *fairness*, and *ingroup* words more often than
conservative ministers who used *authority* and *purity* words more
often. Although conservative ministers were expected to use more
*ingroup* words based on political ideology and previous research, an
examination of the way liberal ministers used *ingroup* words revealed a
tendency for the use of *ingroup* words to glorify rebellion and promote
independence (i.e., the opposite direction from *ingroup* definitions).
Effect sizes indicated relatively sizable difference between liberal and
conservative sermons with Cohen's *d* values ranging from 0.56 to 1.27.

In contrast to this approach, @Sagi2014 explored how moral words were
used paired with other co-occurring concepts using Latent Semantic
Analysis (LSA). They examined three different moral issues in different
contexts to piece out specific moral words and their collocates. First,
they looked at how moral words were used in relation to the World Trade
Center compared to the Empire State Building in the New York Times from
1987-2007. After 9-11, the number of moral words associated with the
World Trade Center increased, specifically *harm* words from the MFD.
Second, they considered the changes in how moral words were paired with
mosque used in blogs as a response to the debate of building a mosque
near Ground Zero following 9-11. They found words from the MFD were used
more often with mosque during the main debate and then the co-occurrence
decreased afterwards. Lastly, they examined moral language tied to the
abortion debate in Congress. Republicans used more moral language
overall; more specifically, Republicans tended to use more words
associated with the *purity* foundation; while Democrats used more words
associated with the *fairness* foundation.

@@

MFT has received criticism on the grounds that its assumptions regarding
moral intuitions have little empirical basis. @Suhler2011 criticized the
content and taxonomy of the five foundations and question whether or not
the foundations are sufficiently distinct as to stand as their own
foundation. Likewise, @Gray2015 argues that the measurement of the moral
foundations confounds morality with other constructs (such as weirdness)
and the moral foundations lack statistical and conceptual distinction.
@Schein2015 argues that harm-based morality is the most parsimonious and
logical explanation for MFT findings.

Beyond the critiques of the theory itself, many have also provided
strong critiques of the measurement using MFD. Conducting two close
conceptual replication studies using religious sermons and six extension
studies using U.S. Congressional speeches, party platforms, State of the
Union addresses, media outlets, and political organization texts,
@Frimer2020 found that the liberal-conservative differences found in the
original study replicated in only 30% of cases and the effect sizes were
over 30 times smaller than reported in @Graham2009. In a meta-analysis,
the theorized differences were only found for authority and purity
(e.g., conservatives used more authority and purity words). Loyalty was
contrary to theoretical predictions (e.g., used more by liberals), and
no differences were found for harm and fairness. Additionally,
@Frimer2020 tested the theory under the most forgiving conditions using
not only the original MFD but also the revised MFD2 as well using log
transformed scores and raw scores. Even under the best circumstances
(log transformed MFD2 scores), only two of the foundations showed
theorized differences but with very small effects. Other studies have
also found conflicting findings which fail to support the basic
assertions of MFT. Exploring the political Twitterverse, @Sterling2018
found differences in the MFD depend on political sophistication. For
example, harm words were used more by liberals only if they had low
levels of political sophistication; at high levels of political
sophistication, conservatives used more harm words. While the MFD has
not been extensively used (at least in published research), studies
which have use the MFD to test partisan differences have found mixed
results in terms of supporting MFT [@Clifford2013; @Frimer2015;
@Sagi2014]. Like @Frimer2020, we attempt to give the MFD the best chance
for success by incorporating a weighting method to address the
skewedness and sparsity of the linguistic data.

The most successful attempts to validate a linguistic measure of moral
foundations have used more complex methods (compared to the original
dictionary-based approach). These attempts generally fall into two
categories: human annotations and semantic representations. @Hopp2021
represents the first approach where instead of relying on 'experts', a
large sample of lay people were crowdsourced to manually annotate
document for content relevant to each foundation. While
resource-intensive, this method better captures how people may encounter
and judge moral issues in everyday life leading to a more ecologically
valid dictionary. @Garten2016 represents the second categories which
instead relies on data-driven semantic analysis. Using shallow neural
net models like word2vec [@mikolov_efficient_2013], moral foundations
were measured using semantic similarity vectors rather than simple word
counts. Here, we try to augment the simpler approach to add more context
to see if the MFD liberal-conservative differences can be replicated.

First, we use the MFQ and MFD in a multi-trait multi-method approach to
determine if the dictionary can be improved and mapped into the
theoretical MFT traits. This study examined the measurement overlap
between these two sources, as well as the latent traits after
controlling for the overlap. Using the results from this study, we
examined two conceptual replications of @Frimer2020. We test the
liberal-conservative difference proposed by MFT using partisan media
content and the MFD. We first test differences in a general news corpus
compared liberal sources (National Public Radio [NPR] and the New York
Times [NYT]) to conservative sources (Fox News and Breitbart). Second,
we examined news about specific political events: Brett Kavanaugh's U.S.
Supreme Court nomination and the U.S. Government Shutdown in 2018-2019.
In the second study, the news sources were expanded to cover NYT, NPR,
Slate, Huffington Post, and Politico on the liberal side and Fox News,
Breitbart, The Rush Limbaugh Show, The Blaze, and Sean Hannity on the
conservative side. To address potential issues of measurement with the
MFD, the MFD was combined with the information produced in Study 1 and
weighted by valence to create weighted percentages to better specify
endorsement. Like @Frimer2020, results were judged based on the
direction of the liberal-conservative difference compared MFT
predictions and the size of the effect compared to past studies.

@@

Each of the foundation pairs were listed together (i.e., *harm/care*,
*fairness/reciprocity,* etc.) with a space for participants to list
their free association concepts. After the free association task,
participants were then given the 15-items from the moral relevance
section of the MFQ as described in the introduction. Last, participants
were asked to denote their political orientation from 1 *conservative*
to 10 *liberal*, as well as which political party they associated with:
Democrat, Republican, and Independent. The survey was delivered through
Qualtrics, and participants were recruited through the online
participant management system for the university (SONA). Each
participant signed an online consent form at the beginning of the study
and was given participation credit at the end of the study.

@@

Participant data was first spell checked and screening for inaccurate
responses. Participants who did not write more than fifty words in
response to a given prompt were excluded (*n*  = 
`r nrow(study1.2DF) - nrow(finalwords)`). One missing datapoint was
estimated using the *mice* library from *R* for a missing MFQ question
[@buuren2011], and all other missing data was present in the
demographics sections, which were not filled in. The MFQ data were
screened for outliers using Mahalanobis distance, and `r nrow(finalwords) - nrow(noout1.2)` outliers were
found at $\chi^2_{p<.001}$(`r ncol(allcolumns[ , -c(1,32:42)])`)  = 
`r printnum(cutoff1.2)`. These data were excluded leading to a final
sample size of `r nrow(noout1.2)`. Data were screened for assumptions
described in Study 1.1 and were found to be satisfactory.

@@

##adding forms/conjugations of words to original_mfd
#harm/care
original_mfd[c(27:95),1] <- c("abusive","abuser","abused","abusing",
                             "sympathetic","sympathies","damaged",
                             "damaging","attacked","attacking","attacker",
                             "attacks","attackers", "benefits", "benefitted",
                             "benefitting", "cares", "cared", "caring", "crushes",
                             "crushing", "crushed", "dangerous", "defends", "defending",
                             "defended", "detroys", "destroyed", "destroying",
                             "fights", "fought", "fighting", "guards", "guarded",
                             "guarding", "harms", "harmed", "harming", "hurts",
                             "hurting", "kills", "killed", "killing", "preserves",
                             "preserved", "preserving", "protects", "protected",
                             "protecting", "protection", "ruins", "ruined", "ruining",
                             "safely", "safer", "shelters", "sheltered", "sheltering",
                             "spurns", "spurned", "spurning", "stomps", "stomped",
                             "stomping", "suffers", "suffered", "suffering",
                             "violence", "warring")

#fairness/reciprocity
original_mfd[c(20:64),2] <- c("balances", "balanced", "balancing","biased",
                             "biases", "discriminates", "discriminated",
                             "discriminating", "discrimination", "equals",
                             "equaled", "equaling", "equates", "equated",
                             "equating", "evens", "evened", "evening",
                             "excludes", "excluded", "excluding", "fairness",
                             "favors", "favored", "favoring", "honesty",
                             "impariality", "justice", "justifies", "justified",
                             "justifying", "prefers", "preferred", "preferring",
                             "prejudices", "prejudiced", "prejudicing", "reasons",
                             "reasoned", "reasoning", "rights", "tolerates",
                             "tolerated", "tolerating", "toleration")

#ingroup/loyalty
original_mfd[c(16:47),3] <- c("collects", "collected", "collecting", "collective",
                             "communities", "deceives", "deceived", "deceiving",
                             "deception", "deceptions", "deserts", "deserted",
                             "deserting", "desertion", "families", "fellows",
                             "foreigners", "groups", "grouped", "grouping",
                             "indviduals", "indvidualize", "individualized",
                             "individualizing", "members", "nations", "sides",
                             "togetherness", "traits", "unites", "united", "uniting")

#authority/respect
original_mfd[c(31:112),4] <- c("abides", "abided", "abiding", "authorities",
                              "classes", "classed", "classing", "command",
                              "commanded", "commanding", "controls", "controlled",
                              "controlling", "defects", "defected", "defecting",
                              "defers", "deferred", "deferring", "deference",
                              "defies", "defied", "defying", "defiance", "deserts",
                              "deserted", "deserting", "desertion", "duties",
                              "faiths", "fathers", "fathered", "fathering",
                              "honors", "honored", "honoring", "laws", "leads",
                              "leading", "mothers", "mothered", "mothering",
                              "obeys", "obeyed", "obeying", "opposes", "opposed",
                              "opposing", "orders", "ordered", "ordering", "permits",
                              "permitted", "permitting", "positions", "positioned",
                              "positioning", "preserves", "preserved", "preserving",
                              "preservation", "protests", "protested", "protesting",
                              "refuses", "refused", "refusing", "refusal", "respects",
                              "respected", "respecting", "respectful", "reveres",
                              "revered", "revering", "reverence", "serves", "served",
                              "serving", "traditions", "traditional", "traits")

#purity/sanctity
original_mfd[c(21:60),5] <- c("abstains", "abstained", "abstaining", "abstinence",
                             "adulteries", "adulterous", "adulterer", "adulterers",
                             "churches", "cleans", "cleaned", "cleaning", "cleanse",
                             "cleanliness", "dirty", "diseases", "diseased", 
                             "disgusts", "disgusted", "disgusting", "grossness",
                             "innocence", "modesty", "preserves", "preserved",
                             "preserving", "preservation", "promiscuity",
                             "promiscuities", "purity", "rights", "ruins", "ruined",
                             "ruining", "sacredness", "sickness", "sicknesses",
                             "sins", "wholeness", "wholesome")

@@

apa_table(mtmmdata.summary, 
          note = "Mean percents for dictionaries, mean total subscores for the MFQ.",
          caption = "Mean Percent Scores and MFQ Scores for Foundation Areas",
          col.names = c("Foundation Area", "Dictionary/MFQ", "M", "SD"),
          format = "latex",
          escape = F,
          align = c("l", "l", "c", "c"),
          font_size = 'footnotesize',
          placement = "h"
          )
```

@@

Fit indices from the MTMM Model 1 step are shown in Table \@ref(tab:mtmm-model1-table). The complete parameter estimates from these models are saved online for full inspection due to their length. The fit indices from these models indicate that all models showed good fit. The original and complete MFD models showed equal fit ($\Delta$ CFA <= .01), and both models were better than the reduced MFD model. We then inspected the loadings to determine the adequacy of the loadings for the MFQ and MFD. These results are shown in Table \@ref(tab:mtmm-model1-table2). All but one question for the MFQ load adequately (>.30) on each of their moral foundation areas. None of the MFD areas load onto their moral foundation area, as all loadings are close to zero (and non-significant, see online). The excellent model fit was likely due to the structure of the MFQ, as the MFD results do not show relation to their foundation area. We did not run further models, as the goal was to show that Model 1 was the best model with appropriate loadings for foundations.

>>>>

abstract: |
  Partisan differences and diviseness have become an increasing hot topic in psychological research. Many theories have been proposed to explain these differences and divisions including Moral Foundations Theory. The current research seeks determine if a linguistic measure of Moral Foundations, the Moral Foundations Dictionary (MFD) can unveil these theoretical partisan differences. In Study 1, we explored the MFD-Moral Foundations Questionnaire's measurement overlap using a multi-trait multi-method model and determined if current predictive shortcomings of the MFD can be improved with additional linguistic content. Next, we examined two replication extensions of previous work assessing the differences in political articles on moral differences using the MFD. Through web scraping, we extracted articles from popular partisan news sources' websites, calculated MFD word frequencies, and identified words' respective valences. This process attempts to uncover news outlets' positive or negative endorsements of certain moral dimensions concomitant with a particular ideology. In Study 2, we gathered political articles from four sources. In Study 3, we expanded their number of sources to 10 and analyzed articles that pertain to two specific topics: the 2018 confirmation hearings of U.S. Supreme Court Justice Brett Kavanaugh and the partial U.S. Government Shutdown of 2018-2019. Results from all studies indicate that the MFD may not consistently show exptected partisan differences, shedding doubt on the usefulness of the MFD.

@@

The field of moral psychology has a long history with modern moral
psychology beginning in the late 1960s with Lawrence Kohlberg's theory
of moral development [@Kohlberg1977]. Since then, Kohlberg's theory has
been highly criticized with numerous theories and hypotheses proposed to
replace it. One of these theories is Moral Foundations Theory (MFT)
proposed by Haidt and colleagues [@haidt2004; @Haidt2007]. While the
theory itself has been widely criticized and debated, the goal of the
current work is to tackle a related measurement issue, the Moral
Foundations Dictionary [MFD, @Graham2009]. @Graham2009 developed the
MFD - a lexicon-based linguistic measure of MFT - arguing that moral
concerns and the propositions of MFT could be measured and tested with
the words people use. Here we first review evidence and criticisms
against this assumption. We then provide our own test of the MFD using a statistical validity assessment and partisan news articles.

@@

Haidt and Joseph [-@haidt2004] and Haidt and Graham [-@Haidt2007]
formulated MFT as a method by which to capture the entirety of humans'
moral domain. They argued that older theories of moral psychology were
focused primarily on issues of justice, fairness, and caring, which are
individually focused foundations of morality that align with the beliefs
of political liberals. In other words, moral psychology ignored the
valid moral foundations of conservatives. MFT holds that people's moral
domain can be mapped by quantifying their endorsement of five moral
foundations: harm/care, fairness/reciprocity, ingroup/loyalty,
authority/respect, and purity/sanctity [@Haidt2007]. Thus, they
proposed that the addition of these two additional domains could
adequately capture the broader spectrum of morality. Throughout the rest of the article, we will use the first word from each of these continuums to denote the foundation area.

Measurement of MFT has traditionally focused on two options: 1) the
Moral Foundations Questionnaire (MFQ) - a typical Likert-type survey
instrument that measures moral relevance of each foundation to an
individual and the moral judgment of various scenarios depicting each
foundation, and 2) the Moral Foundations Dictionary (MFD) which was
designed to capture language's role in moral reasoning and justification
as used in speech and text [@Graham2009; @Graham2011]. The MFQ is a
30-item scale with 15 items on the moral relevance and judgment
scenarios. The 15 moral relevance items are equally divided among the
five foundations and examine how relevant a condition is to making a
moral judgment on a scale of 1 (*not at all relevant*) to 6 (*extremely
relevant*). These relevance items include examples such as: "Whether or
not someone used violence (harm)," "Whether or not someone was denied
his or her rights (fairness)," "Whether or not someone showed a lack
of loyalty (ingroup)," "Whether or not an action caused chaos or
disorder (authority)," and "Whether or not someone did something
disgusting (purity)." The moral judgments items are also equally
divided between the foundations and ask on a six-point scale how much
one agrees with each of the statements. These judgment items include:
"One of the worst things a person can do is hurt a defenseless animal
(harm)," "Justice is the most important requirement of a society
(fairness)," "I am proud of my country's history (ingroup)," "Men
and women each have different roles to play in society (authority),"
and "Chastity is an important and valuable virtue (purity)."

The internal consistency of the MFQ from @Graham2011 was $\alpha$ = .73
averaged across subscales with a range of $\alpha$ = .65-.84. Across six
studies, the MFQ was found to have an average Cronbach's alpha of .63
for harm, .64 for fairness, .56 for ingroup, .59 for authority,
and .71 for purity [@Federico2013; @Graham2009; @Graham2012;
@Weber2013]. Test-retest reliability was *r* = .68-.82 using a sample of
123 college students. Confirmatory factor analysis supported a
well-fitted five-factor model (harm, fairness,
ingroup, authority, and purity) over two,
individual (harm and fairness) versus group (ingroup, authority,
and purity) foundations, or three, autonomy (harm, fairness),
community (ingroup, authority), and divinity (purity) ethics,
foundations factor model. The five-factor structure also fit for
non-Western samples, thus, providing evidence of the MFQ
generalizability. Convergent validity was supported with correlations on
other measures of morality [@Graham2011].

The MFD is composed of 259 words, with around 50 words assigned to each
of the five foundations. @Graham2009 created a preliminary list of words
that they believed would be associated with the five foundations. Words
such as *war* and *peace* should indicate a greater concern with harm
foundation whereas words such as *homeland* and *terrorism* should
indicate a greater concern with the ingroup foundation. The other
foundation dictionaries include *equal* and *justice* (fairness),
*honor* and *protest* (authority), and *holy* and *sin* (purity). To
validate the word sets, @Graham2009 examined the frequency of MFD words
in liberal and conservative sermons using the Linguistic Inquiry and
Word Count [LIWC, @Pennebaker2007] computer program. They found liberal
ministers used harm, fairness, and ingroup words more often than
conservative ministers who used authority and purity words more
often. Although conservative ministers were expected to use more
ingroup words based on political ideology and previous research, an
examination of the way liberal ministers used ingroup words revealed a
tendency for the use of ingroup words to glorify rebellion and promote
independence (i.e., the opposite direction from ingroup definitions).
Effect sizes indicated relatively sizable difference between liberal and
conservative sermons with Cohen's $d_s$ values ranging from 0.56 to 1.27.

In contrast to this approach, @Sagi2014 explored how moral words were
used paired with other co-occurring concepts using Latent Semantic
Analysis (LSA). They examined three different moral issues in different
contexts to piece out specific moral words and their collocates. First,
they looked at how moral words were used in relation to the World Trade
Center compared to the Empire State Building in the New York Times from
1987-2007. After 9-11, the number of moral words associated with the
World Trade Center increased, specifically harm words from the MFD.
Second, they considered the changes in how moral words were paired with
mosque used in blogs as a response to the debate of building a mosque
near Ground Zero following 9-11. They found words from the MFD were used
more often with mosque during the main debate and then the co-occurrence
decreased afterwards. Lastly, they examined moral language tied to the
abortion debate in Congress. Republicans used more moral language
overall; more specifically, Republicans tended to use more words
associated with the purity foundation; while Democrats used more words
associated with the fairness foundation.

@@

MFT has received criticism on the grounds that its assumptions regarding
moral intuitions have little empirical basis. @Suhler2011 criticized the
content and taxonomy of the five foundations and question whether or not
the foundations are sufficiently distinct as to stand as their own
foundation. Likewise, @Gray2015 argued that the measurement of the moral
foundations confounds morality with other constructs (such as weirdness)
and that the moral foundations lack statistical and conceptual distinction.
@Schein2015 argued that harm-based morality is the most parsimonious and
logical explanation for MFT findings.

Beyond the critiques of the theory itself, many have also provided
strong critiques of the measurement using MFD. Conducting two close
conceptual replication studies using religious sermons and six extension
studies using U.S. Congressional speeches, party platforms, State of the
Union addresses, media outlets, and political organization texts,
@Frimer2020 found that the liberal-conservative differences found in the
original study replicated in only 30% of cases and the effect sizes were
over 30 times smaller than reported in @Graham2009. In a meta-analysis,
the theorized differences were only found for authority and purity
(e.g., conservatives used more authority and purity words). Loyalty was
contrary to theoretical predictions (e.g., used more by liberals), and
no differences were found for harm and fairness. Additionally,
@Frimer2020 tested the theory under the most forgiving conditions using
not only the original MFD but also the revised MFD2 as well using log
transformed scores and raw scores. Even under the best circumstances
(log transformed MFD2 scores), only two of the foundations showed
theorized differences but with very small effects.

Other studies have
also found conflicting findings which fail to support the basic
assertions of MFT. Exploring the political Twitterverse, @Sterling2018
found differences in the MFD depend on political sophistication. For
example, harm words were used more by liberals only if they had low
levels of political sophistication; at high levels of political
sophistication, conservatives used more harm words. While the MFD has
not been extensively used (at least in published research), studies
which have use the MFD to test partisan differences have found mixed
results in terms of supporting MFT [@Clifford2013; @Frimer2015;
@Sagi2014]. Like @Frimer2020, we attempt to give the MFD the best chance
for success by both increasing linguistic content and incorporating a weighting method to address the
skewedness and sparsity of the linguistic data.

The most successful attempts to validate a linguistic measure of moral
foundations have used more complex methods (compared to the original
dictionary-based approach). These attempts generally fall into two
categories: human annotations and semantic representations. @Hopp2021
assessed the first approach where instead of relying on 'experts', a
large sample of lay people were crowdsourced to manually annotate
document for content relevant to each foundation. While
resource-intensive, this method better captures how people may encounter
and judge moral issues in everyday life leading to a more ecologically
valid dictionary. @Garten2016 examined the second category which
instead relies on data-driven semantic analysis. Using shallow neural
net models like word2vec [@mikolov_efficient_2013], moral foundations
were measured using semantic similarity vectors rather than simple word
counts. Here, we try to augment the simpler approach to add more context
to see if the MFD liberal-conservative differences can be replicated.

First, we use the MFQ and MFD in a multi-trait multi-method approach to
determine if the dictionary can be improved and mapped into the
theoretical MFT traits. This study examined the measurement overlap
between these two sources, as well as the latent traits after
controlling for method-related measurement. Using the results from this study, we
examined two conceptual replications of @Frimer2020. We test the
liberal-conservative difference proposed by MFT using partisan media
content and the MFD. We first test differences in a general news corpus
compared liberal sources (National Public Radio [NPR] and the New York
Times [NYT]) to conservative sources (Fox News and Breitbart). Second,
we examined news about specific political events: Brett Kavanaugh's U.S.
Supreme Court nomination and the U.S. Government Shutdown in 2018-2019.
In the second study, the news sources were expanded to cover NYT, NPR,
Slate, Huffington Post, and Politico on the liberal side and Fox News,
Breitbart, The Rush Limbaugh Show, The Blaze, and Sean Hannity on the
conservative side. To address potential issues of measurement with the
MFD, the MFD was combined with the information produced in Study 1 and
weighted by valence to create weighted percentages to better specify
endorsement. Like @Frimer2020, results were judged based on the
direction of the liberal-conservative difference compared MFT
predictions and the size of the effect compared to past studies.

@@

Each of the foundation pairs were listed together (i.e., harm,
fairness, etc.) with a space for participants to list
their free association concepts. After the free association task,
participants were then given the 15-items from the moral relevance
section of the MFQ as described in the introduction. Last, participants
were asked to denote their political orientation from 1 *conservative*
to 10 *liberal*, as well as which political party they associated with:
Democrat, Republican, and Independent. The survey was delivered through
Qualtrics, and participants were recruited through the online
participant management system for the university (SONA). Each
participant signed an online consent form at the beginning of the study
and was given participation credit at the end of the study.

@@

Participants who did not write more than fifty words in
response to a given prompt were excluded (*n*  = 
`r nrow(study1.2DF) - nrow(finalwords)`). One missing datapoint was
estimated using the *mice* library from *R* for a missing MFQ question
[@buuren2011], and all other missing data was present in the
demographics sections, which were not filled in. The MFQ data were
screened for outliers using Mahalanobis distance, and `r nrow(finalwords) - nrow(noout1.2)` outliers were
found at $\chi^2_{p<.001}$(`r ncol(allcolumns[ , -c(1,32:42)])`)  = 
`r printnum(cutoff1.2)`. These data were excluded leading to a final
sample size of `r nrow(noout1.2)`. Data were screened for assumptions
described in Study 1.1 and were found to be satisfactory.

@@

##adding forms/conjugations of words to original_mfd
#Harm
original_mfd[c(27:95),1] <- c("abusive","abuser","abused","abusing",
                             "sympathetic","sympathies","damaged",
                             "damaging","attacked","attacking","attacker",
                             "attacks","attackers", "benefits", "benefitted",
                             "benefitting", "cares", "cared", "caring", "crushes",
                             "crushing", "crushed", "dangerous", "defends", "defending",
                             "defended", "detroys", "destroyed", "destroying",
                             "fights", "fought", "fighting", "guards", "guarded",
                             "guarding", "harms", "harmed", "harming", "hurts",
                             "hurting", "kills", "killed", "killing", "preserves",
                             "preserved", "preserving", "protects", "protected",
                             "protecting", "protection", "ruins", "ruined", "ruining",
                             "safely", "safer", "shelters", "sheltered", "sheltering",
                             "spurns", "spurned", "spurning", "stomps", "stomped",
                             "stomping", "suffers", "suffered", "suffering",
                             "violence", "warring")

#Fairness
original_mfd[c(20:64),2] <- c("balances", "balanced", "balancing","biased",
                             "biases", "discriminates", "discriminated",
                             "discriminating", "discrimination", "equals",
                             "equaled", "equaling", "equates", "equated",
                             "equating", "evens", "evened", "evening",
                             "excludes", "excluded", "excluding", "fairness",
                             "favors", "favored", "favoring", "honesty",
                             "impariality", "justice", "justifies", "justified",
                             "justifying", "prefers", "preferred", "preferring",
                             "prejudices", "prejudiced", "prejudicing", "reasons",
                             "reasoned", "reasoning", "rights", "tolerates",
                             "tolerated", "tolerating", "toleration")

#Ingroup
original_mfd[c(16:47),3] <- c("collects", "collected", "collecting", "collective",
                             "communities", "deceives", "deceived", "deceiving",
                             "deception", "deceptions", "deserts", "deserted",
                             "deserting", "desertion", "families", "fellows",
                             "foreigners", "groups", "grouped", "grouping",
                             "indviduals", "indvidualize", "individualized",
                             "individualizing", "members", "nations", "sides",
                             "togetherness", "traits", "unites", "united", "uniting")

#Authority
original_mfd[c(31:112),4] <- c("abides", "abided", "abiding", "authorities",
                              "classes", "classed", "classing", "command",
                              "commanded", "commanding", "controls", "controlled",
                              "controlling", "defects", "defected", "defecting",
                              "defers", "deferred", "deferring", "deference",
                              "defies", "defied", "defying", "defiance", "deserts",
                              "deserted", "deserting", "desertion", "duties",
                              "faiths", "fathers", "fathered", "fathering",
                              "honors", "honored", "honoring", "laws", "leads",
                              "leading", "mothers", "mothered", "mothering",
                              "obeys", "obeyed", "obeying", "opposes", "opposed",
                              "opposing", "orders", "ordered", "ordering", "permits",
                              "permitted", "permitting", "positions", "positioned",
                              "positioning", "preserves", "preserved", "preserving",
                              "preservation", "protests", "protested", "protesting",
                              "refuses", "refused", "refusing", "refusal", "respects",
                              "respected", "respecting", "respectful", "reveres",
                              "revered", "revering", "reverence", "serves", "served",
                              "serving", "traditions", "traditional", "traits")

#Purity
original_mfd[c(21:60),5] <- c("abstains", "abstained", "abstaining", "abstinence",
                             "adulteries", "adulterous", "adulterer", "adulterers",
                             "churches", "cleans", "cleaned", "cleaning", "cleanse",
                             "cleanliness", "dirty", "diseases", "diseased", 
                             "disgusts", "disgusted", "disgusting", "grossness",
                             "innocence", "modesty", "preserves", "preserved",
                             "preserving", "preservation", "promiscuity",
                             "promiscuities", "purity", "rights", "ruins", "ruined",
                             "ruining", "sacredness", "sickness", "sicknesses",
                             "sins", "wholeness", "wholesome")

@@

apa_table(mtmmdata.summary, 
          note = "Mean percents for dictionaries, mean total subscores for the MFQ.",
          caption = "Mean Percent Scores and MFQ Scores for Foundation Areas",
          col.names = c("Foundation Area", "Dictionary/MFQ", "M", "SD"),
          format = "latex",
          escape = F,
          align = c("l", "l", "c", "c"),
          font_size = 'footnotesize',
          placement = "ht"
          )
```

@@

Fit indices from the MTMM Model 1 step are shown in Table \@ref(tab:mtmm-model1-table). The complete parameter estimates from these models are saved online for full inspection due to their length. The fit indices from these models indicate that all models showed good fit. The original and complete MFD models showed equal fit ($\Delta$ CFI <= .01), and both models were better than the reduced MFD model. We then inspected the loadings to determine the adequacy of the loadings for the MFQ and MFD. These results are shown in Table \@ref(tab:mtmm-model1-table2). All but one question for the MFQ load adequately (>.30; *denied rights*) on each of their moral foundation areas. None of the MFD areas load onto their moral foundation area, as all loadings are close to zero (and non-significant, see online). The excellent model fit was likely due to the structure of the MFQ, as the MFD results do not show relation to their foundation area. We did not run further models, as the goal was to show that Model 1 was the best model with appropriate loadings for foundations.

```{r mtmm-model1-table, results = 'asis'}
apa_table(
  bind_rows(
  model1.orig.fit.results, model1.all.fit.results, model1.reduce.fit.results
) %>% 
  mutate(Model = c("Original", "Complete", "Reduced")) %>% 
  select(Model, chisq, npar, rmsea, srmr, cfi, tli, AIC)
  , 
          caption = "MTMM Fit Indices for Model 1 MTMM",
          col.names = c("Model", "Chi-Sq", "df", "RMSEA", "SRMR", "CFI", "TLI", "AIC"),
          format = "latex",
          escape = F,
          align = c("l", "l", "c", "c"),
          font_size = 'footnotesize',
          placement = "ht"
          )
```
